A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-05 paper.tex

\section*{SARS-CoV-2 ARGs: a case study}
% What is SARS-CoV-2 and why is it interesting here? What's the dataset?
% What are its basic evolutionary properties, summarised for a
% population geneticist?
Recombination via template switching is an important
factor in the evolution of
SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}.
With multiple deeply divergent lineages circulating globally,
coinfection (and the potential for recombination)
is relatively likely~\citep{bal2022detection}.
The COVID-19 pandemic led to collection of whole-genome viral data
at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
currently holding over 15 million sequences [FIXME]. Detection
and characterisation of recombination is of profound interest
for [reasons XYZ].

% Summary of sc2ts paper. We inferred some big ole ARGs, with an
% unprecedented degree of detail about recombination (outside, I guess,
% of agri datasets). Summary of what's previously known about sc2 recomb.
\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
on GISAID data, one containing 1.27 million sequences sampled up to
June 30, 2021 (the ``Wide ARG'', and the second
(more sparsely sampled) consisting of 657K sequences sampled up to June 30,
2022 (the ``Long ARG''). Recombination is rare in both ARGs, with $<0.3\%$
of the nodes being inferred to be recombinants (and quite likely
a substantial number of false positives caused by poor data quality).
Nonetheless, this provides thousands of automatically detected
recombination events, the characterisation of which is usually a manual
and labour intensive process. The recombinations events detected
and fully characterised in these large ARGs fit well with previous
which exhaustively investigate specific
recombinants~\citep{jackson2021generation}.
[Not sure what I'm trying to say here;
what does a PopGen person who knows very little about SARS2 need to
know here to given them context?]

% What's the point of this case study? To emphasise that we
% need to systematically represent topological uncertainty in ARGs.
The pandemic scale SARS-CoV-2 dataset is an interesting case study
for a number of reasons.
Firstly, fundamental properties of the data underscore the
need for topological uncertainty through polytomies
and stacked recombination [more concise way of saying this?].
The scale of sequencing effort at the height
of the pandemic means that there were at times tens of thousands
of sequences collected in a single day. This, combined with
the slow rate of mutation, means that we will often have hundreds
of identical sequences collected on the same day.
This makes polytomies an unescapable reality, and a
fully-resolved binary tree is clearly false precision in this case.
At the other extreme, ``unary'' nodes are also a common and
necessary property of such densely sampled data [explain and
give numbers?]
Even with this exquisite level of detail we are
often unable to fully time-resolve recombination events. While
in some cases a recombination event with two parental lineages
can be quite precisely located in space and time (e.g. XA was in Wales
around [when]), in others three or more quite distinct parental
lineages indicate multiple unobserved recombinations.
[JK: I'm just making stuff up here: I'm assuming that there's an
example of a complex X lineages that people haven't figured the
exact history of. Just seeing if this makes a good narrative
before filling in the details.]
While this
may be seen as a limitation of the \texttt{sc2ts} method,
concerted efforts to disentangle the origins of [some example
of a complicated X lineage] have not succeeded in shedding
much extra light, and truthfully representing the uncertainty
around these events seems a more useful approach than
 arbitrary resolution. Encoding the dataset as an eARG,
with a specific event for each recombination breakpoint,
would require such false precision.

An interesting property of the SARS-CoV-2 dataset is that
it highlights the uncertainty in the genomic position of
recombination breakpoints. [Some discussion about why
we can never know where the breakpoint really happened, and
some follow-up to note that this is even more acute in
standard popgen datasets where we're thinking about ancient
recombs].
This is a limitation of the
gARG encoding proposed here, which requires that we specify
exact intervals of inheritance and therefore precise
recombination breakpoints.
An approach in which we can systematically reason about the
uncertainty of inheritance intervals as they propogate
through an ARG
would be an interesting avenue for future research.

@@

% The necessity of complete precision about all recombination
% events under the eARG encoding has been
% conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
% For example, \citet{brandt2021evaluation} define a ``full'' ARG as
% ``a structure that encodes all
% coalescence and recombination events''. While there are certainly cases
% when inferring such fully precise ARGs is useful, for example when
% studying the details of a small number of recombination
% events~\citep{rasmussen2022espalier}, it is worth considering
% whether sufficient information exists to support such inferences
% and indeed whether this precision is required, or useful,
% for downstream applications.

% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
% (equivalent to the ``bricked tree sequence'' of
% \cite{nowbandegani2023extremely},
% % TODO look this up. I expect it's basically the same??
% and the block-based ARG data structure of
% \citep{palamara2016argon}),
% here we consider the unique sets of
% samples that coalesce at a node in the ARG over a particular
% genome interval, and the fundamental limitations they place
% on ARG inference.

% [FIXME find somewhere natural for this. The last clause is too vague - what
% are the implications?]
% Alternative gARG encodings split transmitted intervals and associate each with
% a new node, resulting in a single genome being represented by a group of
% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for

@@

Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination,
but for which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.

@@

>>>>

@@

@@

% Note: haven't said anything about the coalescent here. Do we need to?
The advantages of a model-free representation that naturally
incorporates uncertainty are well illustrated by the
ARGs inferred for SARS-CoV-2 by~\cite{zhan2023towards}.
The COVID-19 pandemic led to collection of whole-genome viral data
at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
holding over 15 million sequences as of [FIXME].
Recombination, while rare, is an important
factor in the evolution of
SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}
with recombinants potentially gaining fitness by combining
mutations from deeply diverged lineages [cite].
\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
on GISAID data, one containing 1.27 million sequences sampled up to
June 30, 2021 and the other (more sparsely sampled) consisting of
657K sequences sampled up to June 30, 2022.
Sequencing effort at the height of the pandemic produced
tens of thousands of sequences collected in a single day,
with hundreds being identical.
Polytomies are therefore and unescapable reality, and a
fully-resolved binary tree would clearly be false precision.
Because of the longitudinal data collection, there is no
requirement that samples must be leaf nodes, and samples
will frequently descend from other (older) samples.
Unary nodes (a node in a local tree with exactly one child)
are also a common feature, when a sample is inferred to have
exactly one direct descendant.
Despite the abundance of data and the rarity of recombination,
the parental lineages cannot be precisely identified
in many cases. For example, many are inferred to have three
or more parental lineages (in agreement with the
analysis resulting in the Pango designation of the recombinant
lineage); in most cases, these will indicate two or more
independent recombination events.
Faithfully representing the uncertainty
around these events seems a more useful approach than
the arbitrary resolution which would be required,
if encoded as an eARG.

% An interesting property of the SARS-CoV-2 dataset is that
% it highlights the uncertainty in the genomic position of
% recombination breakpoints. [Some discussion about why
% we can never know where the breakpoint really happened, and
% some follow-up to note that this is even more acute in
% standard popgen datasets where we're thinking about ancient
% recombs].
% This is a limitation of the
% gARG encoding proposed here, which requires that we specify
% exact intervals of inheritance and therefore precise
% recombination breakpoints.
% An approach in which we can systematically reason about the
% uncertainty of inheritance intervals as they propogate
% through an ARG
% would be an interesting avenue for future research.

@@

[TODO finish up with a positive note about interchange, and make the point
clearly that tskit is an ARG library. Splice in the stuff about haplotype
blocks and ARGON format here somewhere --- basically this is a derived product
of the gARG encoding, so while there are some computation advantages it's not
fundamendal to the actual representation of the thing]

% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
% (equivalent to the ``bricked tree sequence'' of
% \cite{nowbandegani2023extremely},
% % TODO look this up. I expect it's basically the same??
% and the block-based ARG data structure of
% \citep{palamara2016argon}),
% here we consider the unique sets of
% samples that coalesce at a node in the ARG over a particular
% genome interval, and the fundamental limitations they place
% on ARG inference.
% [FIXME find somewhere natural for this. The last clause is too vague - what
% are the implications?]
% Alternative gARG encodings split transmitted intervals and associate each with
% a new node, resulting in a single genome being represented by a group of
% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
